Growth model analysis 3spp WB

### 2019-03-14 MS_Analysis_3spp_4rivers_Patterns_from 2spp.Rmd
added speciesRiver model

# by species (bkt, bnt), by cohort, by river gridded
ggplot(vB %>% filter(cohort %in% 2000:2013, species %in% c('bkt', 'bnt', 'ats')), aes(ydayCumul,cf, group = species, color = (species))) +
  geom_point(size = 0.5, alpha = 0.1) +
 # geom_smooth(span = 0.1, size = 1.5, se = F) +
  geom_smooth(method='gam', alpha=0.1, formula=y~s(x, k = kValue), se = F) +
  facet_grid(riverOrdered ~ cohort) +
  theme_publication()

@@

#######################

# 1. Do we need a RE on year in s()? YES
mod0a <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr'), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod1)

mod0 <- gam(observedLength ~ year + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod1)

AIC(mod0,mod0a) %>% rownames_to_column() %>% arrange(AIC)

# 2. Best structure for year,species,river
mod1 <- gam(observedLength ~ species + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod2 <- gam(observedLength ~ river + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

@@

# 2-way models
mod3 <- gam(observedLength ~ year * species + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod4 <- gam(observedLength ~ year * riverOrdered * riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod4)

mod5 <- gam(observedLength ~ species * riverOrdered * riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod4)

# mod5 is best
AIC(mod0,mod1,mod2,mod3,mod4,mod5) %>% rownames_to_column() %>% arrange(AIC)

# 3-way models
mod6 <- gam(observedLength ~ year * species * riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

# https://www.fromthebottomoftheheap.net/2017/12/14/difference-splines-ii/
# calculates differences from overall spline for each year
# a lot slower
vB$yearOrdered <- factor(vB$year, levels = 2000:2013)
mod6a <- gam(observedLength ~ year * species * riverOrdered + s(ydayCumul, bs = 'cr', by = (yearOrdered)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

# try with river instead of riverOrdered
mod6b <- gam(observedLength ~ year * species * river + s(ydayCumul, bs = 'cr', by = (yearOrdered)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod6b)

# 3-way is best
AIC(mod1,mod2,mod3,mod4,mod5,mod6,mod6a) %>% rownames_to_column() %>% arrange(AIC)

# Any additive 3-way better? NO
mod7 <- gam(observedLength ~ year + species * riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

mod8 <- gam(observedLength ~ year * species + riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

mod9 <- gam(observedLength ~ year + species + riverOrdered + s(ydayCumul, bs = 'cr', by = (year)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

@@

summary(mod6)
plot(mod6)

hist(resid(mod6))
gam.check(mod6)

>>>>

# by species (bkt, bnt), by cohort, by river gridded
ggplot(vB %>% filter(cohort %in% 2000:2013, species %in% c('bkt', 'bnt', 'ats')), aes(ydayCumul,observedLength, group = species, color = (species))) +
  geom_point(size = 0.5, alpha = 0.1) +
 # geom_smooth(span = 0.1, size = 1.5, se = F) +
  geom_smooth(method='gam', alpha=0.1, formula=y~s(x, k = kValue), se = F) +
  facet_grid(riverOrdered ~ cohort) +
  theme_publication()

@@

#######################
vB$yearFactor <- factor(vB$year)
vB$yearOrdered <- factor(vB$year, levels = 2000:2013, ordered = T)
vB$yearOrderedSpecies <- factor(paste(vB$yearOrdered,vB$species, sep = '_'), ordered = T)

vB$cohortFactor <- factor(vB$cohort)
vB$cohortOrdered <- factor(vB$cohort, levels = 2000:2013, ordered = T)
vB$cohortOrderedSpecies <- factor(paste(vB$cohortOrdered,vB$species, sep = '_'), ordered = T)

vB$speciesOrdered <- factor(vB$species, levels = c('bkt','bnt'), ordered = T)
vB$cohortSpeciesRiver <- factor(paste(vB$cohortOrdered,vB$speciesOrdered,vB$riverOrdered, sep = '_'), ordered = T)

vB$speciesRiver <- factor(paste(vB$speciesOrdered,vB$riverOrdered, sep = '_'), ordered = T)

# 1. Do we need separate splines on cohort in s()? YES, cohortOrdered is best
# 
mod0_0 <- bam(observedLength ~ cohort + s(ydayCumul, bs = 'cr'), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod0_1 <- bam(observedLength ~ cohort + s(ydayCumul, bs = 'cr', by = (cohort)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod0_2 <- bam(observedLength ~ cohort + s(ydayCumul, bs = 'cr', by = (cohortFactor)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod0_3 <- bam(observedLength ~ 
                cohort + 
                s(ydayCumul, bs = 'cr') + 
                s(ydayCumul, bs = 'cr', by = (cohortOrdered)), 
              data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod0_4 <- bam(observedLength ~ cohort + 
                s(ydayCumul, bs = 'cr') + 
                s(ydayCumul, bs = 'cr', by = (cohortOrderedSpecies)),
                data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

#mod0_4 is best
AIC(mod0_0,mod0_1,mod0_2,mod0_3,mod0_4) %>% rownames_to_column() %>% arrange(AIC)

# 2. structure of smooths
# with ordered factor, need to include an overall smooth, http://www.sfs.uni-tuebingen.de/~jvanrij/Tutorial/GAMM.html 
mod0_4a <- gam(observedLength ~ cohort + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = (cohortOrderedSpecies)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod0_4b <- gam(observedLength ~ cohortOrdered + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = (cohortOrderedSpecies)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

# mod0_3b is best, but we want to estimate a trend as in mod0_3a
# add + s(ydayCumul, bs = 'cr') to models
AIC(mod0_4,mod0_4a,mod0_4b) %>% rownames_to_column() %>% arrange(AIC)

# 3. gam vs. bam for this big dataset? bam() is MUCH faster!
system.time( mod0_4a_1 <- gam(observedLength ~ cohort + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = (cohortOrderedSpecies)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML") )
# 316 sec
summary(mod0_3a_1)

system.time( mod0_4a_2 <- bam(observedLength ~ cohort + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = (cohortOrderedSpecies)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML") )
# 8 sec
summary(mod0_4a_2)

# 4. bs = 'cr', vs default 'tp'. AIC for 'cr' is ~ 300 lower. Stick with 'cr'
mod0_4a_2a <- bam(observedLength ~ cohort + s(ydayCumul) + s(ydayCumul, by = (cohortOrderedSpecies)), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

AIC(mod0_4a_2,mod0_4a_2a)

# 5. Include Individual as a RE??
# Taking a long time, maybe run overnight
vB$tagIndexFactor <- factor(vB$tagIndex)
system.time( mod0_3a_3 <- bam(observedLength ~ year + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = (yearOrdered)) + s(ydayCumul, tagIndexFactor, bs="fs", m=1), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML") )
# sec
compareML(mod0_3a_3,mod0_3a_2)

# 2. Best structure for year,species,river
mod0 <- bam(observedLength ~ cohort + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = cohortOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod1 <- bam(observedLength ~ species + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = cohortOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod2 <- bam(observedLength ~ river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = cohortOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

@@

# 2-way models
mod3 <- bam(observedLength ~ cohort * species + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = cohortOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod4 <- bam(observedLength ~ cohort * river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = cohortOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

mod5 <- bam(observedLength ~ species * river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = cohortOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

# mod4 is best
AIC(mod0,mod1,mod2,mod3,mod4,mod5) %>% rownames_to_column() %>% arrange(AIC)

# 3-way models
mod6 <- bam(observedLength ~ cohort * species * river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = cohortOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod6)

mod6a <- bam(observedLength ~ cohort * species * river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = cohortOrderedSpecies), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod6a)

mod6b <- bam(observedLength ~ cohort * river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = cohortOrderedSpecies), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod6b)

#mod6a is best
AIC(mod6,mod6a,mod6b) %>% rownames_to_column() %>% arrange(AIC)

# Any additive 3-way better? NO
mod7 <- bam(observedLength ~ cohort + species * river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = cohortOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

mod8 <- bam(observedLength ~ cohort * species + river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = cohortOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

mod9 <- bam(observedLength ~ cohort + species + river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr', by = cohortOrdered), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')))

@@

summary(mod6)
plot(mod6a)

hist(resid(mod6a))
gam.check(mod6a)

plot(mod6a, shade = TRUE, pages = 2,  seWithMean = TRUE)

#################################
# Predictions
# https://www.fromthebottomoftheheap.net/2017/10/10/difference-splines-i/

mod6a <- bam(observedLength ~ cohort * species * river + s(ydayCumul, bs = 'cr') + s(ydayCumul, bs = 'cr'), data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")

predGrid <- expand.grid(cohortOrderedSpecies = unique(mod6a$model$cohortOrderedSpecies), 
                        river = mod6a$xlevels$river,  
                        ydayCumul = seq(250,1000, by = 30)) %>%
            mutate(cohort = as.numeric(substr(cohortOrderedSpecies,1,4)),
                   species = substr(cohortOrderedSpecies,6,8))

pred <- predict(mod6a, newdata = predGrid)#, type = 'lpmatrix')

predGrid2 <- bind_cols(predGrid,as.data.frame(pred))

ggplot(predGrid2, aes(ydayCumul,pred, color = species)) +
  geom_point() + 
  geom_line() +
  facet_grid(river ~ cohort)

mod6csr <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = cohortSpeciesRiver), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod6csr)
plot(mod6csr)

predGrid <- expand.grid(cohortSpeciesRiver = unique(mod6csr$model$cohortSpeciesRiver),
                        ydayCumul = seq(250,1000, by = 30)) %>%
                 mutate(cohort = as.numeric(substr(cohortSpeciesRiver,1,4)),
                        species = substr(cohortSpeciesRiver,6,8),
                        river = substring(cohortSpeciesRiver,10))

pred <- predict(mod6csr, newdata = predGrid)#, type = 'lpmatrix')

predGrid2 <- bind_cols(predGrid,as.data.frame(pred))

ggplot(predGrid2, aes(ydayCumul,pred, color = species)) +
  geom_point() + 
  geom_line() +
  ylim(0,300) +
  facet_grid(river ~ cohort)

mod6sr <- bam(observedLength ~ cohort * species * river + 
               s(ydayCumul, bs = 'cr') +
               s(ydayCumul, bs = 'cr', by = speciesRiver), 
             data = vB %>% filter(cohort %in% 2000:2013,species %in% c('bkt','bnt')), method = "REML")
summary(mod6sr)

predGrid <- expand.grid(speciesRiver = unique(mod6sr$model$speciesRiver),
                        cohort = 2000:2013,
                        ydayCumul = seq(250,1000, by = 30)) %>%
                 mutate(species = substr(speciesRiver,1,3),
                        river = substring(speciesRiver,5))

pred <- predict(mod6sr, newdata = predGrid)#, type = 'lpmatrix')

predGrid2 <- bind_cols(predGrid,as.data.frame(pred))

ggplot(predGrid2, aes(ydayCumul,pred, color = species)) +
  geom_point() + 
  geom_line() +
  facet_grid(river ~ cohort)
